% --------------------------------------------------------------
% METADATA
% Notes by: Scott Schoeller (sschoellerSTEM)
% Author: Rabin, M.O.
% Proceedings Symposia Applied Math
% Year: 1967
% --------------------------------------------------------------

\renewcommand{\thesubsection}{\Roman{subsection}}

\section{MATHEMATICAL THEORY OF AUTOMATA}

\subsection{FINITE AUTOMATA}
word on $\Sigma$
for a finite sequence $x=\sigma_{0}\sigma_{1}...\sigma_{n-1}, \sigma_{i} \in \Sigma$\\
The length $l(x)$ is the no. of elements in \textit{x}\\
$\Lambda$ is the empty word.\\
"$\Sigma^{*}$ is a free semigroup under" concatenation\\
$\sigma \in \Sigma$ are "free generators" of this semi-group\\
For events (or subsets) $A \subseteq \Sigma^{*}, B \subseteq \Sigma^{*}$\\
$A^{*} = A^{0} \cup A ...$\\

\textbf{DEFINITION 1}:
"A \textit{finite automata (f.a.)} over $\Sigma$ is a system $\mathfrak{U} = < S, M, s_0, F>$ 
where \textit{S} is a finite set (the set of \textit{states}), $M: S \times \Sigma \to S$ (the \textit{table of transitions} of $\mathfrak{U}$) 
$s_0 \in S$ ($s_0$ is the \textit{initial state}), $F \subseteq S$ (\textit{F} is the set of \textit{designated} final states)."\\

\textit{M} "can be uniquely extended to" $M^{*}: S \times \Sigma^{*} \to S$; $M^{*}(s, \Lambda) = s, s \in S$\\
$M^{*}(s, x\sigma) = M(M^{*}(s,x), \sigma, s \in S, \sigma \in \Sigma, x \in \Sigma^{*}$\\

\textit{M} is the state-transition function with respect to  $\mathfrak{U}$.\\
In a similar manner, $M^{*}$ describes state-transitions under input \textit{words} $x \in \Sigma^{*}$.\\

\textbf{DEFINITION 2}: 
"The set $T(\mathfrak{U}) = \{ x | x \in \Sigma^{*}, M^{*}(s_0, x) \in F \}$ \\
is the \textit{set} (event) \textit{defined} by $\mathfrak{U}$. A set $A \subseteq \Sigma^{*}$ is called a \textit{regular event} if for some finite automaton $\mathfrak{U}, A = T(\mathfrak{U})$."

\subsubsection{F.a. mappings}
Little distinction between automata in DEFINITION 1 and those that "yield a mapping from input sequences to output sequences".\\
Associate "state $s \in F$ the output 1 and with each $s \in S - F$ the output 0".\\
Associate $\mathfrak{U}$ the mapping $T:\Sigma^{*} \to \{ 0, 1 \}^{*}$ s.t. $T(\sigma_{0} \sigma_{1}...\sigma_{n-1})$,\\
Where for $ i \in [0, n-1], \tau_i = 1$ "if $M^{*}(s_0, \sigma_0 ... \sigma_i) \in F$ and and $r_i = 0$ otherwise."\\
 \newpage
Generalizations possible:\\
1. Partitioning \textit{S} into disjoint "union of \textit{k} sets", where $S= F_0 \cup ... \cup F_{k-1}$; mapping $T:\Sigma^{*} \to { \{ 0, 1, ..., k-1 \} }^{*}$
into words from a \textit{k}-letter alphabet.\\
2. Output of $F_i, i \in,[0, k-1]$ as a word, $w_i$ on alphabet $\Omega$.  Mapping is $T:\Sigma^{*} \to \Omega^{*}$ ("not even length preserving")\\
3. Output as function $f(s, \sigma)$ with respect to the current state and current input of $\mathfrak{U}$

\subsubsection{Nonderministic automata}
$\mathfrak{U}$ in state \textit{s} with input $\sigma$; can go into any one of a number of states $s' \in S'$, where $S \subset S$ is a set depending on $s$ and $\sigma$.\\
As a result of DEFINITION 1: $M: S \times \Sigma \to P(S)$; "$P(S)$ is the power set of $S$ and $s_0$ is replaced by a set $S_0 \subseteq S$ of initial states."\\
If $x=\sigma_0 \sigma_1 ... \sigma_{n-1}$, the sequence of states ($s_0, s_1, ..., s_n$) is \textbf{compatible} w/ x "if $s_0 \in S_0$" and for $i \in [0, n-1]$.\\
$\mathfrak{U}$ accepts $x$ if for some sequence of states through $s_n$ is compatible w/ $x$, $s_n \in F$.\\
The set $T(\mathfrak{U})$ consists of all words accepted by $\mathfrak{U}$.\\
\\
THEOREM 1: "For every nonderministic automata $\mathfrak{U}$ there exists a f.a. $\mathfrak{B}$" s.t. "$T(\mathfrak{U}) = T(\mathfrak{B})$. 
If $\mathfrak{U}$ has n states, then $\mathfrak{B}$" has less than "$2^n$ states." % correction to wording made here


\subsubsection{Regular expressions and events}
Let $Y_1, Y_2, ...$ be vars "ranging over subsets of $\Sigma^{*}$  The set of regular terms in $Y_1, Y_2, ...$ , $\mathbf{R}$ "is the smallest set satisfying the conditions":\\
1. $Y_n \in \mathbf{R}$, where $n \in [1, \infty)$\\
2. if $R_1 \in \mathbf{R}, R_2 \in \mathbf{R}$ then $R_1 \cup R_2 \in \mathbf{R}, (R_1 R_2) \in \mathbf{R}, R_{1}^{*} \in \mathbf{R}$\\
Every element of $\textbf{R}$ is called a regular expression $R(\sigma_1, ..., \sigma_n)$, which is a singleton set.\\
A regular expression a way of expressing subsets of $\Sigma^*$\\
THEOREM 2. 
"A set $T \subseteq \Sigma^*$ is f.a. definable (regular)" iff $\exists R(\sigma_1, ..., \sigma_n)$ where $\sigma_i \in \Sigma, i \in [1,k]$ s.t. $T = R(\sigma_1, ..., \sigma_k)$\\
Every regular event is representable by a regular expression.\\  
Two regular terms $R(Y_1, ..., Y_n)$ and $Q(Y_1, ..., Y_n)$ are called equivalent if $\forall A_1 \subseteq \Sigma^*, ...,  \forall A_n \subseteq \Sigma^*$,\\
$R(A_1, ..., A_n) = Q(A_1, ..., A_n)$\\
\\
The "equivlance of regular terms is effectively solvable."\\
If fixed letters of $\Sigma$ are part of "term formulation", $Y_n \in \mathbf{R^{'}}, \sigma \in \mathbf{R^{'}}, n = 1,2, ... \sigma \in \Sigma$.\\
Regular expressions yield f.a. definable sets.

\subsubsection{Algebraization of f.a.}
$E$ is an equiv. relation on $\Sigma^{*}$ is a right-invariant if $xEy$ implies $xzEyz \forall x, y, z \in \Sigma^{*}$\\
left-invariant(?)\\
congruence on $\Sigma^{*}$: given a set T that is a a subset of $\Sigma^{*}$, one can define two relations...\\
\\
THEOREM 3.\\
$T \subseteq \Sigma^{*}$ is regular iff it's "the union of equiv. classes of a right-invariant relation $E$ (on $\Sigma^{*}$) with finite index."  $T$ is regular iff "$E_{T}$ has a finite index."\\
$\exists \mathfrak{U}$ "with index($E_{T}$) states" s.t. $T = T(\mathfrak{U})$. "No automaton with fewer than index($E_{T}$) states defines $T$."\\
\\
THEOREM 3 "is very useful in show that certain sets $T \subseteq \Sigma^{*}$ are or are not regular."\\
\\
THEOREM 4."A set $T \subseteq \Sigma^{*}$ is regular" iff "it is a union of equivalence classes of a congruence $\equiv$ (on $\Sigma^{*}$ with finite index."  "...$T$ is regular" iff "index($\equiv_{T}$) is finite."\\
"Given a congruence relation" on $\Sigma^{*}$, the partitioned set $\Sigma^{*} / \equiv$ of equivalence classes of $\Sigma^{*}$ with respect to $\equiv$, into semi-group so that mapping $\phi_0 : x \to [x]_{\equiv}$  of ea. $x \in \Sigma^{*}$ into equiv. class is a homomorphism of $\Sigma^{*}$ onto $\Sigma^{*} / \equiv$.\\
Conversely, to ea. homomorphism $\phi_1 : \Sigma^{*} \to M$ of $\Sigma^{*}$ onto semi-group, $M$ - corresponding congruence $\equiv$ def by $x \equiv y$ iff $\phi_1 (x) = \phi_1 (y)$.\\
\\
THEOREM 4 (restated).\\
"A set $T \subseteq \Sigma^{*}$ is regular" iff $\exists M, H \subseteq M$ and a homomorphism $\phi : \Sigma^{*} \to M$ s.t. $T = \phi^{-1} (H)$, where M is a finite semigroup.\\
\\
Associate with ea. $\sigma \in \Sigma$ a function $f_{\sigma}$.  The system $<A, a_0, f_{\sigma}>_{\sigma \in \Sigma}$ is an \textit{algebra of type} $\Sigma$ if $a_0 \in A$

\subsubsection{Decision problems and algorithms}
Skipped.

\subsubsection{Finite automata and infinite sequences}
A f.a. $\mathfrak{U}$ accepting a word, $x = \sigma_{0} \sigma_{1}...\sigma_{n}...$, $x$ is a denumerable seq. of letters $\sigma_n \in \Sigma$.\\
Let $\mathfrak{U}$ be a NFA. A denumerable seq. of states ($s_{0}, s_{1}...$) "is called \textit{compatible} with $x$ if $s_{0} \in S_{0}$\\ 
and $S_{n+1} \in M(s_{n}, \sigma_{n}),0 \le n < \infty$."\\

DEFINITION 3. "$\mathfrak{U}$ \textit{accepts} $x$ if for $s$" (a sequence), "$\{ n | S_n \in F \}$ is infinite."  "Set of all denumerable" seq. "accepted by $\mathfrak{U}$ will be denoted by $T_{\infty}(\mathfrak{U})$"\\

THEOREM 6 (B{\"u}chi)."If $H_1 \subseteq \Sigma_{\infty}$ and $H_2 \Sigma_{\infty}$" are f.a. definable, so are the union, intersection of $H_1$ and $H_2$ as well as $\Sigma_{\infty} - H_1$

\subsection{PROBALISTIC AUTOMATA}
Not relevant to problem.

\subsection{TREE-AUTOMATA}
Tree automata have a tree for input.\\
A version of tree automata can be applied to context-free languages [Mezei and Wright, "Generalized Algol-like Languages"].\\
$\mathfrak{U} = <S, M, S_0, F >$ f.a over $\Sigma$; $x = \sigma_{0}\sigma_{1}\sigma_{2} \in \Sigma$.  Terminal node has no value.  Except for the initial state $s_0$, elements of $F$ not indexed.\\
\\
$M: S \times S \times \Sigma \to S$ (note: $M$ earlier defined as transition table).  Inputs for $\mathfrak{U}$ finite binary trees, nodes val. by $\Sigma$.

\subsubsection{Binary trees}
Paper defines trees "as sets of words on alphabet $\Omega = \{ 0,1 \}$."  Let $x,y \in \Omega^{*}, x$ is "...the prefix of $y$ if some $z \in \Omega^{*}, y = xz$."  
"...Subset $H \subseteq \Omega^{*}$ is \textit{prefix closed} if $y \in H$ and $x$ prefix of $y$ imply $x \in H$.   
"The \textit{prefix closure} of" $G \subseteq \Omega^{*}$ "is the smallest prefix closed set $H$..." s.t. $G \subseteq H$.  All "prefix closed set[s]" have the empty word as a member.\\

"A (binary) tree is a prefix-closed subset $H \subseteq \Omega^{*}$."  ."..Elements $x \in H$" are nodes.  "The node $\Lambda \in H$ is the" tree's root.  (why use the empty word symbol?)\\
If $x \in H$ and $x_0 \notin H$ and $x_1 \notin H$, $x$ is a terminal node.\\
$H$ is \textit{frontiered} if $\forall \ x \in H$ ($x$ is non-terminal) we have $x_0 \in H$ and $x_1 \in H$. $Ft(H)$ is "the set of all terminal nodes of $H$."

\subsubsection{Tree-automata}
"A $\Sigma$-valued tree is a pair $(v, H)$, where $H$ is a finite frontiered tree; $v: (H-Fr(H)) \to \Sigma$. Each non-terminal node $x \in H$ is assigned a symbol $v(x) \in \Sigma$.\\
"The set of all valued trees" is denoted by $V(\Sigma, \Omega)$.\\
\\
DEFINITION 9: "A \textit{finite tree automaton} (f.t.a) is a system $\mathfrak{U} = < S,M_t, s_0, F >$." Symbols $S, s_0, F$ in DEFINITION 1.  $M_t : S \times S \times \Sigma \to S$\\
DEFINITION 10: "The f.t.a. $\mathfrak{U}_t$ is said to accept $(v, H) \in V(\Sigma, \Omega)$, if $\exists \ \phi : H \to S$ satisfying:"\\
1. "$\phi(x) = s_0$ for $x \in Fr(H)$"\\
2. "if $x \in Fr(H)$ (hence, $x_0, x_1 \in H$) then $\phi(x) = M_t (\phi(x_0) \phi(x_1), v(x))$"\\
3. "$\phi(\Lambda) \in F$"\\
Set of all inputs accepted denoted $T(\mathfrak{U)}$.  $T$ is the set defined by $\mathfrak{U}_t$.  $T \subseteq V(\Sigma, \Omega)$ is f.t.a. definable if for some $\mathfrak{U}_t, T = T(\mathfrak{U}_t)$.\\
Note: The function $\phi$ is unique.\\
non-deterministic f.t.a. -- $\phi : H \to S$ compatible with $(v,H)$, $T(\mathfrak{U}_t)$ defined by a non-deterministic f.t.a.\\
Results that follow from f.a. to finite tree automata:\\
"...Class of all f.t.a. definable subsets $T \subseteq V(\Sigma, \Omega)$ is a boolean algebra." 
"For every non-deterministic f.t.a" $T(\mathfrak{U}_t) \exists$ a deterministic f.t.a. $\mathfrak{B}_t$ s.t. $T(\mathfrak{U}_t) = T(\mathfrak{B}_t)$.\\
"The emptiness and equivalence problems for f.t.a are solvable." (? what is the emptiness problem; does the equivalence problem have to do with the eq. relation material earlier?)

\newpage

\subsubsection{Valued trees and terms}
There is "a one-to-one correspondence between terms and valued trees..."\\
$T_0 = (v_0, H_0)$ and $T_1 = (v_1, H_1)$ are $\Sigma$-valued trees and $\sigma \in \Sigma$.  $H_0$ and $H_1$ are trees, which can be combined: $H = \{ \Lambda \} \cup 0H_0 \cup 1H_1$.\\
\textit{combined valued tree} is denoted $C(\sigma, T_0, T_1)$ is defined to be $(v_0, H)$ where $v(\Lambda) = \sigma, v(0x) = v_0(x)$ for $x \in H_0, v(1x) = v_1(x)$ for $x \in H_1$.\\
"Associate with each $\sigma \in \Sigma$" a binary function (operation) symbol $F_\sigma$ and let $u_0$ be a fixed symbol.  $V_2$, set of all terms, is "...the
 smallest set" s.t. "$u_0 \in V_2$ if $t_0 \in V_2$ and $t_1 \in V_2$ then $F_\sigma(t_0, t_1) \in V$ for $\sigma \in \Sigma$."\\
 $\phi(\Lambda) =  u_0, \phi(C(\sigma, T_0, T_1)) = F_\sigma(\phi(T_0), \phi(T_1)), \sigma \in \Sigma, T_0, T_1 \in V(\Sigma, \Omega)$.\\
"A set $K \subseteq V_2$ of terms is...\textit{f.t.a definable} if for some definable $T \subseteq V(\Sigma, \Omega)$ we have $K = \phi(T)$.\\
If terms $t \in V_2$ are words on the alphabet $\{ F_{\sigma} | \sigma \in \Sigma \}$, definable sets that are not always regular events per DEFINITION 2.\\
"$t$ is often stored in memory in tree (list) form and the evaluation of" $\phi(x)$ for each node done successively, "starting from the terminals for which $\phi(x) = s_0$, ending with $\phi(\Lambda)$."